Peptide retention time prediction

### 2015-11-09 retention.tex
edits from word grammar

% say something about gradient and washing 
Since the elution of the peptides is determined by the proportion of
solvent B present at each time point, the scheme used to add solvent B
is an essential aspect of any RPLC experiment. Currently, the most
commonly used approach is based on {\em linear gradients}, where
the increase in solvent B follows a linear function. Most of the
peptides elute before the percentage of solvent B reaches 40\%, and
thus gradients often go from 2 to 40\% B. A high increase in
percentage B at the end of the run ensures that all the remaining
compounds are washed off from the column.

@@

However, in practice such predictive models are not as straight
forward to design as it might seem. For instance, one important part
of such a model would be the peptides three dimensional
conformation. Although the current structure prediction methods are
quite accurate in predicting three dimensional conformation of short
amino acid chains, such methods will not be equipped with the
flexibility to model the various conditions of a chromatographic
experiment. For instance, many peptides are likely changing their
confirmation as a function of the amount of organic solvent in the
mobile phase, an effect that is likely to have an important
contribution to each peptides retention time. Hence, most efforts to
predict retention time have moved into treating the chromatographic
process as a black box, that is, a system that is only viewed in terms
of its inputs and outputs.

@@

A limitation of index-based methods are that they are optimized for
predicting retention times for certain chromatography setups. Thus,
their prediction ability deteriorates for experimental conditions
diverging from the setup on which the algorithm was optimized
\cite{Spicer2007}. Consequently, more advanced methods such as {\sc
  Elude} \cite{elude1,elude2} estimate retention time indices for each
dataset as hand. In these cases the indices are used to generate
features that are used as an input for final retention time
estimates. We will come back to this in section \ref{sec:ML}.

@@

In a follow-up work, the method was extended by representing each
peptide by 1052 instead of 20 features
\cite{petritis2006improved}. The new descriptors encoded the identity
of each amino acid at each position in a sequence, as well as the
length and the hydrophobic moment of the peptide. The new neural
network was trained on more than 300,000 peptide sequences. To obtain
such a large number of peptides with known retention times, a strategy
to normalize elution times was developed. This allowed pooling of
peptides identified in more than 12,000 shotgun experiments. The
trained algorithm reported better performance for the 1303 peptides
on which it was evaluated, than its predecessor.

@@

% very few studies have documented the chromatographic behavior of
% modified peptides
The study of peptide separation mechanisms in RPLC has been an active
area of research for the past 30 years. Nonetheless, surprisingly
little efforts have been directed toward investigating the
chromatographic behavior of posttranslationally modified peptides. The
few studies available so far have mostly focused on phosphorylated
peptides \cite{Kim2007}, or a couple of artifactual modifications
commonly introduced during sample preparation
\cite{Reimer2011}. Consequently, few retention time prediction methods
have been extended and assessed for such peptides. A majority of the
approaches currently available are modification-specific, many of them
developed specifically for phosphopeptides \cite{Kawakami2005,
  perlova2010}.

@@

It is also worth noting that the current trend of processing huge data
repositories, frequently referred to as big data, have brought into
attention so called deep learning techniques such as belief networks
using restricted multiple layered Boltzmann machines (RBM)
\cite{salakhutdinov2009deep} or deep neural networks. This follows the
observation that the performance of predictors is often enhanced if
one can structure the prediction into multiple layers. The drawback of
such models are that they require large training sets. Given the very
large sets of peptides with retention times that are assembled in
various repositories, such as Peptide Atlas
\cite{desiere2006peptideatlas} and PRIDE \cite{martens2005pride}, such
methods could probably be implemented for retention time prediction as
well.

@@

One way to determine a predictable separation techniques usefulness is
to estimate its ability to remove candidate peptides of any detected
feature. This can be measured by a false positive ratio (FPR), also
known as the method's fall-out. The FPR is of course a function of the
selected thresholds, but if we select a threshold where $\sim$95\% of
the known examples get correctly included, what FPR do we find at such
level?  Current retention time prediction algorithms have an FPR of
about $\frac{1}{5}$, meaning that it can remove about four out of five
candidate peptides for a given feature. This could be compared to the
FPR of about $\frac{1}{200}$ for predictions of high-resolution
iso-electric focusing experiments~\cite{branca2014hirief}. The
retention time prediction has to improve quite a bit to come to level
with the fall-out of isoelectric focusing, however, almost all mass
spectrometry-based proteomics use reversed phase chromatography, while
very few studies are using iso-electric focusing.

>>>>

% say something about gradient and washing 
Since the proportion of solvent B determines the elution of peptides,
the scheme used to add solvent B is an essential aspect of any RPLC
experiment. Currently, the most commonly used approach is based on
{\em linear gradients}, where the increase in solvent B follows a
linear function. Most of the peptides elute before the percentage of
solvent B reaches 40\%, and thus gradients often go from 2 to 40\%
B. A high increase in percentage B at the end of the run ensures that
all the remaining compounds are washed off from the column.

@@

However, in practice such predictive models are not as straight
forward to design, as it might seem. For instance, one important part
of such a model would be the peptides three-dimensional
conformation. Although the current structure prediction methods are
quite accurate in predicting three-dimensional conformation of short
amino acid chains, such methods will not be equipped with the
flexibility to model the various conditions of a chromatographic
experiment. For instance, many peptides are likely changing their
confirmation as a function of the amount of organic solvent in the
mobile phase, an effect that is likely to have an important
contribution to each peptides retention time. Hence, most efforts to
predict retention time have moved into treating the chromatographic
process as a black box, that is, a system that is only viewed in terms
of its inputs and outputs.

@@

A limitation of index-based methods is that they are optimized for
predicting retention times for certain chromatography setups. Thus,
their prediction ability deteriorates for experimental conditions
diverging from the setup on which the algorithm was optimized
\cite{Spicer2007}. Consequently, more advanced methods such as {\sc
  Elude} \cite{elude1,elude2} estimate retention time indices for each
dataset as hand. In these cases the indices are used to generate
features that are used as an input for final retention time
estimates. We will come back to this in section \ref{sec:ML}.

@@

The method was extended by representing each peptide by 1052 instead
of 20 features in a follow-up study~\cite{petritis2006improved}. The
new descriptors encoded the identity of each amino acid at each
position in a sequence, as well as the length and the hydrophobic
moment of the peptide. The new neural network was trained on more than
300,000 peptide sequences. To obtain such a large number of peptides
with known retention times, a strategy to normalize elution times was
developed. This allowed pooling of peptides identified in more than
12,000 shotgun experiments. The trained algorithm reported better
performance for the 1303 peptides on which it was evaluated, than its
predecessor.

@@

% very few studies have documented the chromatographic behavior of
% modified peptides
The study of peptide separation mechanisms in RPLC has been an active
area of research for the past 30 years. Nonetheless, surprisingly
little efforts have been directed toward investigating the
chromatographic behavior of post-translationally modified
peptides. The few studies available so far have mostly focused on
phosphorylated peptides \cite{Kim2007}, or a couple of artifactual
modifications commonly introduced during sample preparation
\cite{Reimer2011}. Consequently, few retention time prediction methods
have been extended and assessed for such peptides. A majority of the
approaches currently available are modification-specific, many of them
developed specifically for phosphopeptides \cite{Kawakami2005,
  perlova2010}.

@@

It is also worth noting that the current trend of processing huge data
repositories, frequently referred to as big data, have brought into
attention so called deep learning techniques such as belief networks
using restricted multiple layered Boltzmann machines (RBM)
\cite{salakhutdinov2009deep} or deep neural networks. This follows the
observation that the performance of predictors is often enhanced if
one can structure the prediction into multiple layers. The drawback of
such models is that they require large training sets. Given the very
large sets of peptides with retention times that are assembled in
various repositories, such as Peptide Atlas
\cite{desiere2006peptideatlas} and PRIDE \cite{martens2005pride}, such
methods could probably be implemented for retention time prediction as
well.

@@

One way to determine a predictable separation technique's usefulness is
to estimate its ability to remove candidate peptides of any detected
feature. This can be measured by a false positive ratio (FPR), also
known as the method's fall-out. The FPR is of course a function of the
selected thresholds, but if we select a threshold where $\sim$95\% of
the known examples get correctly included, what FPR do we find at such
level?  Current retention time prediction algorithms have an FPR of
about $\frac{1}{5}$, meaning that it can remove about four out of five
candidate peptides for a given feature. This could be compared to the
FPR of about $\frac{1}{200}$ for predictions of high-resolution
isoelectric focusing experiments~\cite{branca2014hirief}. The
retention time prediction has to improve quite a bit to come to level
with the fall-out of isoelectric focusing, however, almost all mass
spectrometry-based proteomics use reversed phase chromatography, while
very few studies are using iso-electric focusing.
